---
title: "Hair Time"
author: "LKP & AR"
date: "2026-02-04"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initial note:

Figure 4A is derived from Dataset 1  (n = 4531).
Figures 4B, 4C, 5A, and 5C are based on Dataset 2 (n = 3949), while Figure 5B is generated from Dataset 3 (n = 1450).  
Hairtime (pDLMO) distribution is consistent across all datasets. However, we re-plotted it to maintain a consistent structure across markdown sections and display the distribution of the sample from which each figure originates.

# Figure 4A 

## Distribution - linear 

```{r}
rm(list = ls(all = TRUE))

load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht1.Rda")

suppressMessages(library(ggplot2))
suppressMessages(library(tidyverse))

ht1$hair_time <- ifelse(ht1$hair_time > 13, ht1$hair_time -24, ht1$hair_time)

# plot 

mean = mean(ht1$hair_time)
median = median(ht1$hair_time)
print(paste0("Mean of hairtime: ", mean(ht1$hair_time)))
print(paste0("Median of hairtime: ", median(ht1$hair_time)))
distr <- 
  ggplot() + 
  geom_histogram(data = ht1, aes(x = hair_time),
                 fill = "grey", color = "black", 
                 breaks = seq(-12,12,.5),
                 alpha = .5, linewidth = .1) +
  scale_y_continuous(name = "Distribution (counts)", breaks = seq(0, 900, 150)) +
  scale_x_continuous(limits = c(-12,12),
                     breaks = seq(-12,12,2),
                     labels = c("12:00","14:00", "16:00",
                                "18:00","20:00", "22:00",
                                "00:00",
                                "02:00", "04:00", "06:00",
                                "08:00", "10:00", "12:00"),
                     name = "Predicted DLMO (hours)") +
  geom_vline(xintercept = mean, color = "red", linetype = "dashed") +
  geom_vline(xintercept = median, color = "blue", linetype = "dashed") +
  # add annotation for vlines
  annotate("text", x = mean+2, y = 800, label = "Mean", color = "red", size = 4) +
  annotate("text", x = median-2, y = 800, label = "Median", color = "blue", size = 4) +
  theme_bw()+
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        legend.position = "none",
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.line =  element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -3),
        axis.title.y = element_text(vjust = 3))
distr
```

## Distribution - Circular

```{r}
library(circular)
library(ggplot2)

# Convert hair_time from hours to radians (for circular calculation)
ht_circular <- circular(ht1$hair_time * (2 * pi / 24), units = "radians", template = "clock24")

# Calculate circular mean and median
circular_mean_rad <- mean.circular(ht_circular)
circular_median_rad <- median.circular(ht_circular)

# Convert circular mean and median back to hours
circular_mean_hour <- as.numeric(circular_mean_rad) * (24 / (2 * pi))
circular_median_hour <- as.numeric(circular_median_rad) * (24 / (2 * pi))

# Adjust the plot code to include circular mean and median
distr <- 
  ggplot() + 
  geom_histogram(data = ht1, aes(x = hair_time),
                 fill = "#66a1ff", color = "black", 
                 breaks = seq(-12, 12, 0.5),
                 alpha = 0.5, linewidth = 0.1) +
  geom_vline(xintercept = circular_mean_hour, color = "red", linetype = "dashed") +
  geom_vline(xintercept = circular_median_hour, color = "blue", linetype = "dashed") +
  annotate("text", x = circular_mean_hour + 0.5, y = 800, 
           label = "Circular Mean", color = "red", size = 3, vjust = 0, hjust = 0) +
  annotate("text", x = circular_median_hour - 0.5, y = 750, 
           label = "Circular Median", color = "blue", size = 3, vjust = 0, hjust = 1) +
  scale_y_continuous(name = "Distribution (counts)", breaks = seq(0, 900, 150)) +
  scale_x_continuous(limits = c(-12, 12),
                     breaks = seq(-12, 12, 2),
                     labels = c("12:00", "14:00", "16:00",
                                "18:00", "20:00", "22:00",
                                "24:00", "02:00", "04:00", "06:00",
                                "08:00", "10:00", "")) +
  coord_polar(start = pi) +  # Use coord_polar instead of coord_radial
  theme_bw() +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        legend.position = "none",
        plot.margin = margin(0.8, 0.8, 0.8, 0.6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
        axis.line = element_line(linewidth = 0.4),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = 0.1),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(vjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -3),
        axis.title.y = element_text(vjust = 3))

distr

```



# Figure 4B

## Distribution

```{r}
rm(list = ls(all = TRUE)) 
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")

library(ggplot2)
library(tidyverse)
library(dplyr)

# hair time around midnight
ht2$hair_time <- ifelse(ht2$hair_time > 13, ht2$hair_time -24, ht2$hair_time)

ht2$age_cat_label <- cut(ht2$age, breaks = seq(17,91,3), labels = seq(19, 90, 3))
ht2$age_cat_label <- as.numeric(as.character(ht2$age_cat_label))
ht2$age_cat_label[ht2$age_cat_label == 70] <- 68
ht2$age_cat_label[ht2$age_cat_label == 67] <- 68


ht2$age_cat <- cut(ht2$age, seq(17,71, 3))
ht2$age_cat <- as.character(ht2$age_cat)
ht2$age_cat[ht2$age_cat == "(68,71]"] <- "(65,70]"
ht2$age_cat[ht2$age_cat == "(65,68]"] <- "(65,70]"

# get dots 
means <- ht2 %>% group_by(age_cat_label) %>% summarise(mean_age = mean(age),
                                                       se_hair_time = plotrix::std.error(hair_time),
                                                       mean_hair_time = mean(hair_time),
                                                       n = n())

means$x <- as.numeric(as.character(means$age_cat_label))
means <- means %>% rename(hair_time = mean_hair_time)


# first linear histogram to show distribution used to model (ht2)
mean = mean(ht2$hair_time)
median = median(ht2$hair_time)

print(paste0("Mean of hairtime: ", mean(ht2$hair_time)))
print(paste0("Median of hairtime: ", median(ht2$hair_time)))

distr2 <- 
  ggplot(ht2) +
  geom_histogram(data = ht2, aes(x = hair_time),
                 fill = "grey", color = "black", 
                 breaks = seq(-12,12,.5),
                 alpha = .5, linewidth = .1) +
  theme_bw() +
  scale_y_continuous(name = "Distribution (counts)", 
                     breaks = seq(0, 500, 100)) +
  scale_x_continuous(limits = c(-12,12),
                     breaks = seq(-12,12,2),
                     labels = c("12:00","14:00", "16:00",
                                "18:00","20:00", "22:00",
                                "00:00",
                                "02:00", "04:00", "06:00",
                                "08:00", "10:00", "12:00"),
                     name = "Predicted DLMO (hours)") +
  geom_vline(xintercept = mean, color = "red", linetype = "dashed") +
  geom_vline(xintercept = median, color = "blue", linetype = "dashed") +
  # add annotation for vlines
  annotate("text", x = mean+1, y = 800, label = "Mean", color = "red", size = 3) +
  annotate("text", x = median-1, y = 800, label = "Median", color = "blue", size = 3) +
  theme_bw()+
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        legend.position = "none",
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),  # Add this line
        axis.line =  element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -3),
        axis.title.y = element_text(vjust = 3)) 
distr2
```

## Association between variables 

### Tests

```{r}
suppressMessages(library(circular))
suppressMessages(library(CircStats))
suppressMessages(library(dplyr))
suppressMessages(library(Directional))
suppressMessages(library(car))

# Perform correlation test
corr <- cor.test(log(ht2$age), ht2$hair_time)
corr

# Perform t-test
median(ht2$age)
ht2$age_group <- ifelse(ht2$age < median(ht2$age), "young", "old")
leveneTest(hair_time ~ age_group, data = ht2)
# p > 0.05 → variances are similar → you can use var.equal = TRUE
# p ≤ 0.05 → variances differ → use var.equal = FALSE (Welch)
t.test(ht2$hair_time ~ ht2$age_group, var.equal = F)
```

### Sensitivity analysis (circularity of pDLMO): Watson-Williams test (circular) 
```{r}
suppressMessages(library(circular))
suppressMessages(library(CircStats))
suppressMessages(library(dplyr))
suppressMessages(library(Directional))

# Define younger and older groups based on median age
median_age <- median(ht2$age, na.rm = TRUE)
ht2$age_group <- ifelse(ht2$age < median_age, "young", "old")

# Extract numeric vectors for young and old
young <- ht2 %>% filter(age_group == "young") %>% pull(hair_time)
old <- ht2 %>% filter(age_group == "old") %>% pull(hair_time)
all <- ht2  %>% pull(hair_time)

# Convert hours to radians (since 24 hours = 2π radians)
young_rad <- young * (2 * pi / 24)
old_rad <- old * (2 * pi / 24)
all_rad <- all * (2 * pi / 24)

# Ensure all values are within [0, 2π]
young_rad <- (young_rad + 2 * pi) %% (2 * pi)
old_rad <- (old_rad + 2 * pi) %% (2 * pi)
all_rad <- (all_rad + 2 * pi) %% (2 * pi)

# Convert to circular data
young_circ <- circular(young_rad, units = "radians")
old_circ <- circular(old_rad, units = "radians")
all_circ <- circular(all_rad, units = "radians")

# 1. Linear circular correlation
circlin.cor(theta = all_circ, x = ht2$age, rads = TRUE)

# 1. Perform Watson-Williams test for homogeneity of means
ww_test <- watson.williams.test(list(young_circ, old_circ))
print(ww_test)

# 2. If concentration parameters differ, perform Rao's Spacing Test (uniformity)
rao_test <- rao.spacing.test(c(young_circ, old_circ))
print(rao_test)

# 3. Non-parametric alternative (circular)
watson_wheeler_test <- watson.wheeler.test(list(young_circ, old_circ))
print(watson_wheeler_test)

# 4. Convert circular mean differences back to hours for interpretability
young_mean <- mean.circular(young_circ)
old_mean <- mean.circular(old_circ)

# Convert circular means from radians to hours
young_time <- (young_mean * 12 / pi) + 24
old_time <- (old_mean * 12 / pi) + 24

# Calculate the difference in minutes
time_diff_minutes <- (young_time - old_time) * 60
print(paste("Time difference between young and old groups (minutes):", round(time_diff_minutes, 2)))
```
## Models

### Regression (model A)


```{r }

library(splines)
m1 <- lm(hair_time ~  sex + age, data = ht2)
m2 <- lm(hair_time ~  sex + bs(age), data = ht2)

jtools::summ(m1, digits = 3, confint = T)
jtools::summ(m2, digits = 3, confint = T)

```

#### BIC: 

```{r}
print(paste0("BIC for model 1 (age linear): ", BIC(m1)))
print(paste0("BIC for model 2 (age splines): ", BIC(m2)))
print("Selected: model 2")
```

### Model assumptions

```{r, fig.width=8, fig.height=8}
suppressMessages(library(performance))
check_model(m2)
```

### Plot

```{r , warning=FALSE, fig.width=6, fig.height=4}
library(ggplot2)
# get lines ----------------------------------------------------

eff_age <- ggeffects::predict_response(m2, margin = "marginalmeans", terms = "age")
eff_sex <- ggeffects::predict_response(m2, margin = "marginalmeans", terms = "sex")
eff <- ggeffects::predict_response(m2, margin = "marginalmeans", terms = c("age", "sex"))
eff <- eff %>% rename(hair_time = predicted)
eff_age <- eff_age %>% rename(hair_time = predicted)


# get dots ----------------------------------------------------

ht2$age_cat_label <- cut(ht2$age, breaks = seq(17,91,3),
                         labels = seq(19, 90, 3))
ht2$age_cat_label <- as.numeric(as.character(ht2$age_cat_label))
ht2$age_cat_label[ht2$age_cat_label == 70] <- 68
ht2$age_cat_label[ht2$age_cat_label == 67] <- 68


ht2$age_cat <- cut(ht2$age, seq(17,71, 3))

ht2$age_cat <- as.character(ht2$age_cat)
ht2$age_cat[ht2$age_cat == "(68,71]"] <- "(65,70]"
ht2$age_cat[ht2$age_cat == "(65,68]"] <- "(65,70]"

# get dots 
means <- ht2 %>% group_by(age_cat_label) %>% summarise(mean_age = mean(age),
                                                        se_hair_time = plotrix::std.error(hair_time),
                                                        mean_hair_time = mean(hair_time),
                                                        n = n())

means$x <- as.numeric(as.character(means$age_cat_label))
means <- means %>% rename(hair_time = mean_hair_time)

as_distr <- 
  ggplot() +
  geom_point(data = means, aes(x = x, 
                               y = hair_time, 
                               size = n), 
             color = "grey75") +
  geom_errorbar(data = means, aes(x = x, 
                                  ymin = hair_time-se_hair_time,
                                  ymax = hair_time+se_hair_time),
                linewidth = .3, width = .4, color = "grey75") + 
  geom_line(data = eff_age, aes(x = x, y = hair_time)) +
  geom_ribbon(data = eff_age, aes(ymin = conf.low, 
                                  ymax = conf.high,
                                  x = x), alpha = .2) +
  theme_bw() +
  scale_x_continuous(name = "Age", breaks = seq(20, 80, 10)) +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-4,-3.5,-3, -2.5, -2),
                     labels = c("20.00", "20.50",
                                "21.00", "21.50", "22.00"),
                     limits = c(-4, -2)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.line =  element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3),
        legend.position = c(1, 1),
        legend.justification = c(1, 1)) +
  scale_size(name = "Bin sample size", 
             limits = c(80, 500), range = c(1,5),
             breaks = c(100, 200, 300, 500)) 

as_distr
```

### Sensitivity analysis (circularity of pDLMO): hair time using sine and cosine components
     
```{r}
#rm(list = ls(all = TRUE)) 
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")

library(dplyr)
library(ggplot2)
library(splines)
library(ggeffects) # The star of the show
library(plotrix)   # For std.error

# 1. Prepare Circular Data
# Convert time (0-24) to Angle -> Sine/Cosine
ht2$theta <- (ht2$hair_time %% 24) * (2*pi/24)
ht2$sin_t <- sin(ht2$theta)
ht2$cos_t <- cos(ht2$theta)

# 2. Fit the Two Component Models
# We use 'bs(age)' for the curve and 'sex' as a covariate
m_sin <- lm(sin_t ~ bs(age) + sex, data = ht2)
m_cos <- lm(cos_t ~ bs(age) + sex, data = ht2)


# Get Sine predictions
eff_sin <- predict_response(m_sin, terms = "age", margin = "marginalmeans")

# Get Cosine predictions
eff_cos <- predict_response(m_cos, terms = "age", margin = "marginalmeans")


# Helper function to approximate SE from 95% CI
get_se <- function(fit, upper) (upper - fit) / 1.96

# Create a clean dataframe for the simulation
sim_data <- data.frame(
  age = eff_sin$x,
  sin_fit = eff_sin$predicted,
  sin_se  = get_se(eff_sin$predicted, eff_sin$conf.high),
  cos_fit = eff_cos$predicted,
  cos_se  = get_se(eff_cos$predicted, eff_cos$conf.high)
)

# Run the simulation (vectorized for speed)
n_sims <- 1000
results <- list()

for(i in 1:nrow(sim_data)) {
  # 1. Simulate distributions for Sin and Cos
  s_sim <- rnorm(n_sims, sim_data$sin_fit[i], sim_data$sin_se[i])
  c_sim <- rnorm(n_sims, sim_data$cos_fit[i], sim_data$cos_se[i])
  
  # 2. Convert to Time
  angles <- atan2(s_sim, c_sim)
  times  <- (angles %% (2*pi)) * 24/(2*pi)
  
  # 3. Handle Midnight Wrap (centering on Noon)
  # If time < 12, add 24. This makes 01:00 become 25:00
  times_shifted <- ifelse(times < 12, times + 24, times)
  
  # 4. Store stats
  results[[i]] <- data.frame(
    x = sim_data$age[i],
    hair_time = mean(times_shifted),
    conf.low  = quantile(times_shifted, 0.025),
    conf.high = quantile(times_shifted, 0.975)
  )
}

# Bind results into one final prediction frame
eff_final <- do.call(rbind, results)

ht2$age_cat_label <- cut(ht2$age, breaks = seq(17,91,3), labels = seq(19, 90, 3))
ht2$age_cat_label <- as.numeric(as.character(ht2$age_cat_label))

# Fix outliers
ht2$age_cat_label[ht2$age_cat_label == 70] <- 68
ht2$age_cat_label[ht2$age_cat_label == 67] <- 68

# Shift Raw Data for Plotting (Noon to Noon)
ht2$hair_time_shifted <- ifelse(ht2$hair_time < 12, ht2$hair_time + 24, ht2$hair_time)

# Calculate Means
means <- ht2 %>% 
  group_by(age_cat_label) %>% 
  summarise(
    x = mean(age_cat_label),
    mean_hair_time = mean(hair_time_shifted, na.rm=TRUE),
    se_hair_time   = std.error(hair_time_shifted, na.rm=TRUE),
    n = n()
  ) %>% 
  rename(hair_time = mean_hair_time)


ggplot() +
  # 1. Raw Data (Dots & Error Bars)
  geom_errorbar(data = means, aes(x = x, 
                                  ymin = hair_time - se_hair_time,
                                  ymax = hair_time + se_hair_time),
                linewidth = 0.3, width = 0.4, color = "grey75") +
  
  geom_point(data = means, aes(x = x, y = hair_time, size = n), 
             color = "grey75") +
  
  # 2. Model Prediction (Ribbon & Line from our Simulation)
  geom_ribbon(data = eff_final, aes(x = x, ymin = conf.low, ymax = conf.high), 
              fill = "firebrick", alpha = 0.2) +
  
  geom_line(data = eff_final, aes(x = x, y = hair_time), 
            color = "firebrick", linewidth = 1) +
   
  # 3. Styling
  theme_bw() +
  scale_x_continuous(name = "Age", breaks = seq(20, 80, 10)) +
  
  # Y-AXIS SCALE
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(20, 20.5, 21, 21.5, 22), 
                     labels = c("20.00", "20.50", "21.00", "21.50", "22.00")) + # Close parenthesis here!
  
  # ZOOM LIMITS (Must be a separate layer)
  coord_cartesian(ylim = c(20.05, 22.0)) +
  
  theme(
    plot.background = element_rect(fill = "transparent", colour = NA),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    axis.title = element_text(size = 12, face = "bold"),
    legend.position = c(1, 1),
    legend.justification = c(1, 1)
  ) +
  
  scale_size(name = "Bin sample size", range = c(1, 5))
```

### Model A comparison: Spline-based vs Circular pDLMO

```{r}
# ============================================================================
# COMPARING SPLINE-BASED MODEL vs CIRCULAR MODEL
# ============================================================================
# For Model (1): hair_time ~ sex + bs(age)
#
# Linear Model: Single model with hair_time as response
# Circular Model: Two component models (sine and cosine) 
#
# Since circular model uses 2 models, we sum their AIC/BIC values
# ============================================================================

cat("\n=== MODEL (1) COMPARISON: SPLINE-BASED vs CIRCULAR ===\n\n")

# Spline-based Model (m2 is the selected splines model)
cat("SPLINE-BASED MODEL (bs(age)):\n")
cat("  Formula: hair_time ~ sex + bs(age)\n")
cat("  AIC:", AIC(m2), "\n")
cat("  BIC:", BIC(m2), "\n")
cat("  Log-Likelihood:", logLik(m2)[1], "\n")
cat("  R-squared:", summary(m2)$r.squared, "\n")
cat("  Adj R-squared:", summary(m2)$adj.r.squared, "\n")
cat("  N parameters:", length(coef(m2)), "\n\n")

# Circular Model (m_sin and m_cos)
cat("CIRCULAR MODEL:\n")
cat("  Formula (Sin): sin_t ~ sex + bs(age)\n")
cat("  Formula (Cos): cos_t ~ sex + bs(age)\n")
cat("  AIC (Sin):", AIC(m_sin), "\n")
cat("  AIC (Cos):", AIC(m_cos), "\n")
cat("  AIC (Combined):", AIC(m_sin) + AIC(m_cos), "\n")
cat("  BIC (Sin):", BIC(m_sin), "\n")
cat("  BIC (Cos):", BIC(m_cos), "\n")
cat("  BIC (Combined):", BIC(m_sin) + BIC(m_cos), "\n")
cat("  Log-Likelihood (Combined):", logLik(m_sin)[1] + logLik(m_cos)[1], "\n")
cat("  R-squared (Sin):", summary(m_sin)$r.squared, "\n")
cat("  R-squared (Cos):", summary(m_cos)$r.squared, "\n")
cat("  N parameters (total):", length(coef(m_sin)) + length(coef(m_cos)), "\n\n")

# Comparison
cat("COMPARISON:\n")
delta_aic <- (AIC(m_sin) + AIC(m_cos)) - AIC(m2)
delta_bic <- (BIC(m_sin) + BIC(m_cos)) - BIC(m2)
cat("  Delta AIC (Circular - Linear):", delta_aic, "\n")
cat("  Delta BIC (Circular - Linear):", delta_bic, "\n\n")

if(delta_aic < -2) {
  cat("  CONCLUSION (AIC): Circular model is BETTER (Delta AIC < -2)\n")
} else if(delta_aic > 2) {
  cat("  CONCLUSION (AIC): Linear model is BETTER (Delta AIC > 2)\n")
} else {
  cat("  CONCLUSION (AIC): Models are equivalent (|Delta AIC| <= 2)\n")
}

if(delta_bic < 0) {
  cat("  CONCLUSION (BIC): Circular model is BETTER (Delta BIC < 0)\n")
} else {
  cat("  CONCLUSION (BIC): Linear model is BETTER (Delta BIC > 0)\n")
}

cat("\n  NOTE: Lower AIC/BIC indicates better model fit.\n")
cat("  AIC penalizes complexity less than BIC.\n")
cat("  For circular time data, the circular model is theoretically more appropriate,\n")
cat("  but the linear model may be preferred if fit is comparable due to simplicity.\n")
cat("\n=============================================================\n")
```


# Figure 4C

## Distribution

```{r}
#rm(list = ls(all = TRUE)) 
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")

library(ggplot2)
library(tidyverse)
library(dplyr)

# hair time around midnight
ht2$hair_time <- ifelse(ht2$hair_time > 13, ht2$hair_time -24, ht2$hair_time)

ht2$age_cat_label <- cut(ht2$age, breaks = seq(17,91,3), labels = seq(19, 90, 3))
ht2$age_cat_label <- as.numeric(as.character(ht2$age_cat_label))
ht2$age_cat_label[ht2$age_cat_label == 70] <- 68
ht2$age_cat_label[ht2$age_cat_label == 67] <- 68


ht2$age_cat <- cut(ht2$age, seq(17,71, 3))
ht2$age_cat <- as.character(ht2$age_cat)
ht2$age_cat[ht2$age_cat == "(68,71]"] <- "(65,70]"
ht2$age_cat[ht2$age_cat == "(65,68]"] <- "(65,70]"

# get dots 
means <- ht2 %>% group_by(age_cat_label) %>% summarise(mean_age = mean(age),
                                                       se_hair_time = plotrix::std.error(hair_time),
                                                       mean_hair_time = mean(hair_time),
                                                       n = n())

means$x <- as.numeric(as.character(means$age_cat_label))
means <- means %>% rename(hair_time = mean_hair_time)

print(paste0("Mean of hairtime: ", mean(ht2$hair_time)))
print(paste0("Median of hairtime: ", median(ht2$hair_time)))


# first linear histogram to show distribution used to model (ht2)
mean = mean(ht2$hair_time)
median = median(ht2$hair_time)
distr2 <- 
  ggplot(ht2) +
  geom_histogram(data = ht2, aes(x = hair_time),
                 fill = "grey", color = "black", 
                 breaks = seq(-12,12,.5),
                 alpha = .5, linewidth = .1) +
  theme_bw() +
  scale_y_continuous(name = "Distribution (counts)", 
                     breaks = seq(0, 500, 100)) +
  scale_x_continuous(limits = c(-12,12),
                     breaks = seq(-12,12,2),
                     labels = c("12:00","14:00", "16:00",
                                "18:00","20:00", "22:00",
                                "00:00",
                                "02:00", "04:00", "06:00",
                                "08:00", "10:00", "12:00"),
                     name = "Predicted DLMO (hours)") +
  geom_vline(xintercept = mean, color = "red", linetype = "dashed") +
  geom_vline(xintercept = median, color = "blue", linetype = "dashed") +
  # add annotation for vlines
  annotate("text", x = mean+1, y = 800, label = "Mean", color = "red", size = 3) +
  annotate("text", x = median-1, y = 800, label = "Median", color = "blue", size = 3) +
  theme_bw()+
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        legend.position = "none",
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),  # Add this line
        axis.line =  element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -3),
        axis.title.y = element_text(vjust = 3)) 
distr2
```

## Comparing groups 

### Tests

```{r}
library(car)
# Extract numeric vectors for M and F
m <- ht2 %>% filter(sex == "M") %>% pull(hair_time)
f <- ht2 %>% filter(sex == "F") %>% pull(hair_time)
leveneTest(hair_time ~ sex, data = ht2)
# p > 0.05 → variances are similar → you can use var.equal = TRUE
# p ≤ 0.05 → variances differ → use var.equal = FALSE (Welch)
# Perform t-test 
print(t.test(m, f), val.equal=T)
```
### Plot

```{r, warning=FALSE}
library(ggpubr)

# Compute t-test for hair_time between sexes
t_test_result <- t.test(ht2$hair_time ~ ht2$sex, var.equal = TRUE)
p_value_violin <- t_test_result$p.value

# Create violin plot
violin <- ggplot(ht2, aes(x = sex, y = hair_time)) + 
  geom_violin(aes(color = sex, fill = sex)) +
  geom_boxplot(color = "black", alpha = .5, width = .5) +
  theme(legend.position = "none") +
  scale_color_manual(values = c("#6fa8dc", "#d5a6bd")) +
  scale_fill_manual(values = c("#6fa8dc", "#d5a6bd")) +
  xlab("Sex") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-10, -7.5, -5, -2.5, 0, 2.5, 5.0),
                     labels = c("14.00", "16.50", "19.00", "21.50", "00.00", "02.50", "05.00"))  +
  theme_bw() +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line = element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  stat_pvalue_manual(
    data = data.frame(group1 = "M", group2 = "F", 
                      label = sprintf("p = %.3f", p_value_violin)),
    y.position = 5,
    size = 4
  )

violin
```
### Sensitivity analysis (circularity of pDLMO): Watson-Williams test (circular) 
```{r}
# Convert hours to radians (since 24 hours = 2π radians)
m_rad <- m * (2 * pi / 24)
f_rad <- f * (2 * pi / 24)
 
# Ensure all values are within [0, 2π]
m_rad <- (m_rad + 2 * pi) %% (2 * pi)
f_rad <- (f_rad + 2 * pi) %% (2 * pi)
 
# Convert to circular data
m_circ <- circular(m_rad, units = "radians")
f_circ <- circular(f_rad, units = "radians")
 
# Perform Watson-Williams test
watson.williams.test(list(m_circ, f_circ))
 
print((-0.8551704 * 12 / 3.14159265359) + 24)
print((-0.9232520 * 12 / 3.14159265359) + 24)
 
print((((-0.8551704 * 12 / 3.14159265359) + 24)-((-0.9232520 * 12 / 3.14159265359) + 24)) * 60)


```



## Models

### Regression (model A)

```{r}
# Build model
m1 <- lm(hair_time ~ sex + bs(age), data = ht2)
jtools::summ(m1, digits = 3, confint = T)

```

### Model assumptions

```{r, fig.width=8, fig.height=8}
suppressMessages(library(performance))
check_model(m1)

```

### Plot (estimated marginal means) 

```{r, warning=FALSE}

library(emmeans)
library(ggpubr)

# Compute emmeans for pairwise comparisons
emmeans_results <- emmeans(m1, ~ sex)
pairwise_comparison <- contrast(emmeans_results, method = "pairwise")

# Extract p-value
p_value_sex <- summary(pairwise_comparison)$p.value

# Get effect estimates for sex
eff_sex <- ggeffects::predict_response(m1, margin = "marginalmeans", terms = "sex")

# Create mod_sex plot
mod_sex <- ggplot() +
  geom_point(data = eff_sex, aes(x = x, y = predicted, color = x), size = 3) +
  geom_errorbar(data = eff_sex, aes(x = x, ymin = conf.low, ymax = conf.high, color = x), 
                linewidth = .4, width = .05) + 
  theme_bw() +
  xlab("Sex") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-4.00, -3.75, -3.5, -3.25, -3),
                     labels = c("19.00", "20.25", "20.50", "20.75", "21.00"),
                     limits = c(-3.8, -3.2)) +  
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line = element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  scale_color_manual(values = c("#6fa8dc", "#d5a6bd"), name = "Working") +
  scale_fill_manual(values = c("#6fa8dc", "#d5a6bd"), name = "Working") +
  stat_pvalue_manual(
    data = data.frame(x = "F", y = 4.5, group1 = "M", group2 = "F", 
                      label = sprintf("p = %.3f", p_value_sex)),
    y.position = -3.3,
    size = 5
  )

mod_sex
```



### Sensitivity analysis (circularity of pDLMO): hair time using sine and cosine components 

```{r}
#rm(list = ls(all = TRUE)) 
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")

library(dplyr)
library(ggplot2)
library(splines)
library(ggeffects)
library(ggpubr)
library(emmeans)

# hair time around midnight
ht2$hair_time <- ifelse(ht2$hair_time > 13, ht2$hair_time -24, ht2$hair_time)

# 1. Prepare Circular Data
# Convert time (0-24) to Angle -> Sine/Cosine
ht2$theta <- (ht2$hair_time %% 24) * (2*pi/24)
ht2$sin_t <- sin(ht2$theta)
ht2$cos_t <- cos(ht2$theta)

# 2. Fit the Two Component Models
m_sin <- lm(sin_t ~ bs(age) + sex, data = ht2)
m_cos <- lm(cos_t ~ bs(age) + sex, data = ht2)

# Get predictions for sex
eff_sin_sex <- predict_response(m_sin, terms = "sex", margin = "marginalmeans")
eff_cos_sex <- predict_response(m_cos, terms = "sex", margin = "marginalmeans")

# Helper function to approximate SE from 95% CI
get_se <- function(fit, upper) (upper - fit) / 1.96

# Create a clean dataframe for the simulation
sim_data_sex <- data.frame(
  sex = eff_sin_sex$x,
  sin_fit = eff_sin_sex$predicted,
  sin_se  = get_se(eff_sin_sex$predicted, eff_sin_sex$conf.high),
  cos_fit = eff_cos_sex$predicted,
  cos_se  = get_se(eff_cos_sex$predicted, eff_cos_sex$conf.high)
)

# Run the simulation
n_sims <- 1000
results_sex <- list()

for(i in 1:nrow(sim_data_sex)) {
  s_sim <- rnorm(n_sims, sim_data_sex$sin_fit[i], sim_data_sex$sin_se[i])
  c_sim <- rnorm(n_sims, sim_data_sex$cos_fit[i], sim_data_sex$cos_se[i])
  
  angles <- atan2(s_sim, c_sim)
  times  <- (angles %% (2*pi)) * 24/(2*pi)
  times_shifted <- ifelse(times < 12, times + 24, times)
  
  results_sex[[i]] <- data.frame(
    x = sim_data_sex$sex[i],
    hair_time = mean(times_shifted),
    conf.low  = quantile(times_shifted, 0.025),
    conf.high = quantile(times_shifted, 0.975)
  )
}

eff_sex_final <- do.call(rbind, results_sex)

# Compute p-value for circular model
# Test both sin and cos components jointly using Hotelling's T-squared approach
# Or use Wald test on both components
emmeans_sin <- emmeans(m_sin, ~ sex)
emmeans_cos <- emmeans(m_cos, ~ sex)

# Get p-values from both components
pairwise_sin <- contrast(emmeans_sin, method = "pairwise")
pairwise_cos <- contrast(emmeans_cos, method = "pairwise")

# For circular data, we need to test both components jointly
# Simple approach: use the more conservative (larger) p-value
p_value_sin <- summary(pairwise_sin)$p.value
p_value_cos <- summary(pairwise_cos)$p.value

# Use Bonferroni correction or take max (more conservative)
p_value_circular_sex <- max(p_value_sin, p_value_cos)

print(paste("Circular model - Sin component p-value:", p_value_sin))
print(paste("Circular model - Cos component p-value:", p_value_cos))
print(paste("Circular model - Combined p-value:", p_value_circular_sex))

# Create plot
mod_sex_circular <- ggplot() +
  geom_point(data = eff_sex_final, aes(x = x, y = hair_time, color = x), size = 3) +
  geom_errorbar(data = eff_sex_final, aes(x = x, ymin = conf.low, ymax = conf.high, color = x), 
                linewidth = .4, width = .05) + 
  theme_bw() +
  xlab("Sex") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(20.25, 20.5, 20.75),
                     labels = c("20.25", "20.50", "20.75")) +
  coord_cartesian(ylim = c(20.2, 20.8)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line = element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  scale_color_manual(values = c("#6fa8dc", "#d5a6bd"), name = "Sex") +
  scale_fill_manual(values = c("#6fa8dc", "#d5a6bd"), name = "Sex") +
  stat_pvalue_manual(
    data = data.frame(x = "F", y = 4, group1 = "M", group2 = "F", 
                      label = sprintf("p = %.3f", p_value_circular_sex)),
    y.position = 20.75,
    size = 5, tip_length = 0.02
  )


mod_sex_circular
```


### Model A comparison: Spline-based vs Circular pDLMO

```{r}
# ============================================================================
# COMPARING SPLINE-BASED MODEL vs CIRCULAR MODEL - MODEL (2)
# ============================================================================
# For Model (2): hair_time ~ sex + bs(age)
# Focus: Sex effect with age adjustment
# ============================================================================

cat("\n=== MODEL (2) COMPARISON: SPLINE-BASED vs CIRCULAR ===\n\n")

# Spline-based Model (m1)
cat("SPLINE-BASED MODEL (bs(age)):\n")
cat("  Formula: hair_time ~ sex + bs(age)\n")
cat("  AIC:", AIC(m1), "\n")
cat("  BIC:", BIC(m1), "\n")
cat("  Log-Likelihood:", logLik(m1)[1], "\n")
cat("  R-squared:", summary(m1)$r.squared, "\n")
cat("  Adj R-squared:", summary(m1)$adj.r.squared, "\n")
cat("  N parameters:", length(coef(m1)), "\n\n")

# Circular Model (m_sin and m_cos)
cat("CIRCULAR MODEL:\n")
cat("  Formula (Sin): sin_t ~ sex + bs(age)\n")
cat("  Formula (Cos): cos_t ~ sex + bs(age)\n")
cat("  AIC (Sin):", AIC(m_sin), "\n")
cat("  AIC (Cos):", AIC(m_cos), "\n")
cat("  AIC (Combined):", AIC(m_sin) + AIC(m_cos), "\n")
cat("  BIC (Sin):", BIC(m_sin), "\n")
cat("  BIC (Cos):", BIC(m_cos), "\n")
cat("  BIC (Combined):", BIC(m_sin) + BIC(m_cos), "\n")
cat("  Log-Likelihood (Combined):", logLik(m_sin)[1] + logLik(m_cos)[1], "\n")
cat("  R-squared (Sin):", summary(m_sin)$r.squared, "\n")
cat("  R-squared (Cos):", summary(m_cos)$r.squared, "\n")
cat("  N parameters (total):", length(coef(m_sin)) + length(coef(m_cos)), "\n\n")

# Comparison
cat("COMPARISON:\n")
delta_aic <- (AIC(m_sin) + AIC(m_cos)) - AIC(m1)
delta_bic <- (BIC(m_sin) + BIC(m_cos)) - BIC(m1)
cat("  Delta AIC (Circular - Linear):", delta_aic, "\n")
cat("  Delta BIC (Circular - Linear):", delta_bic, "\n\n")

if(delta_aic < -2) {
  cat("  CONCLUSION (AIC): Circular model is BETTER (Delta AIC < -2)\n")
} else if(delta_aic > 2) {
  cat("  CONCLUSION (AIC): Linear model is BETTER (Delta AIC > 2)\n")
} else {
  cat("  CONCLUSION (AIC): Models are equivalent (|Delta AIC| <= 2)\n")
}

if(delta_bic < 0) {
  cat("  CONCLUSION (BIC): Circular model is BETTER (Delta BIC < 0)\n")
} else {
  cat("  CONCLUSION (BIC): Linear model is BETTER (Delta BIC > 0)\n")
}

cat("\n  NOTE: Lower AIC/BIC indicates better model fit.\n")
cat("  The circular model accounts for the periodic nature of time data.\n")
cat("\n=============================================================\n")
```


# Figure 5A

## Distribution

```{r}
# Clear environment
#rm(list = ls(all = TRUE)) 

# Load dataset
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")

# number of NA in ht2$employed
sum(is.na(ht2$employed))
ht2 <- ht2 %>% filter(!is.na(employed))

# Load required libraries
library(ggplot2)
library(tidyverse)
library(dplyr)
library(splines)
library(ggpubr)
library(emmeans)

# Adjust hair_time
ht2$hair_time <- ifelse(ht2$hair_time > 13, ht2$hair_time - 24, ht2$hair_time)



# linear histogram to show distribution used to model
# first linear histogram to show distribution used to model (ht2)
mean = mean(ht2$hair_time)
median = median(ht2$hair_time)

print(paste0("Mean of hairtime: ", mean(ht2$hair_time)))
print(paste0("Median of hairtime: ", median(ht2$hair_time)))

distr2 <- 
  ggplot(ht2) +
  geom_histogram(data = ht2, aes(x = hair_time),
                 fill = "grey", color = "black", 
                 breaks = seq(-12,12,.5),
                 alpha = .5, linewidth = .1) +
  theme_bw() +
  scale_y_continuous(name = "Distribution (counts)", 
                     breaks = seq(0, 500, 100)) +
  scale_x_continuous(limits = c(-12,12),
                     breaks = seq(-12,12,2),
                     labels = c("12:00","14:00", "16:00",
                                "18:00","20:00", "22:00",
                                "00:00",
                                "02:00", "04:00", "06:00",
                                "08:00", "10:00", "12:00"),
                     name = "Hair Time (h)") +
  geom_vline(xintercept = mean, color = "red", linetype = "dashed") +
  geom_vline(xintercept = median, color = "blue", linetype = "dashed") +
  # add annotation for vlines
  annotate("text", x = mean+1, y = 800, label = "Mean", color = "red", size = 3) +
  annotate("text", x = median-1, y = 800, label = "Median", color = "blue", size = 3) +
  theme_bw()+
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        legend.position = "none",
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),  # Add this line
        axis.line =  element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -3),
        axis.title.y = element_text(vjust = 3)) 
distr2
```
## Comparing groups 

### Testing

```{r}
suppressMessages(library(circular))
suppressMessages(library(car))
# Extract numeric vectors for M and F
em <- ht2 %>% filter(employed == "Yes") %>% pull(hair_time)
nem <- ht2 %>% filter(employed == "No") %>% pull(hair_time)
leveneTest(hair_time ~ employed, data = ht2)
# p > 0.05 → variances are similar → you can use var.equal = TRUE
# p ≤ 0.05 → variances differ → use var.equal = FALSE (Welch)
# Perform t-test 
print(t.test(em, nem), var.equal = F)
```

### Plot 

```{r}
# Create violin plot
violin <- ggplot(ht2, aes(x = employed, y = hair_time)) + 
  geom_violin(aes(color = employed, fill = employed)) +
  geom_boxplot(color = "black", alpha = .5, width = .5) +
  theme(legend.position = "none") +
  scale_color_manual(values = c("#6D93B0", "#B08A6D")) +
  scale_fill_manual(values = c("#6D93B0", "#B08A6D")) +
  xlab("Working") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-10, -7.5, -5, -2.5, 0, 2.5, 5),
                     labels = c("14.00", "16.50", "19.00", "21.50", "00.00", "02.50", "05.00"))  +
  theme_bw() +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line =  element_line(linewidth = .4),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  stat_pvalue_manual(
    data = data.frame(group1 = "No", group2 = "Yes",
                      label = sprintf("p = %.3f", p_value_violin)),
    y.position = 5,
    size = 4
  )
violin
```

### Sensitivity analysis (circularity of pDLMO): Watson-Williams test (circular) 

```{r}
# Convert hours to radians (since 24 hours = 2π radians)
em_rad <- em * (2 * pi / 24)
nem_rad <- nem * (2 * pi / 24)
 
# Ensure all values are within [0, 2π]
em_rad <- (em_rad + 2 * pi) %% (2 * pi)
nem_rad <- (nem_rad + 2 * pi) %% (2 * pi)
 
# Convert to circular data
em_circ <- circular(em_rad, units = "radians")
nem_circ <- circular(nem_rad, units = "radians")
 
# Perform Watson-Williams test
watson.williams.test(list(em_circ, nem_circ))
 
# print((-0.8551704 * 12 / 3.14159265359) + 24)
# print((-0.9232520 * 12 / 3.14159265359) + 24)
#  
# print((((-0.8551704 * 12 / 3.14159265359) + 24)-((-0.9232520 * 12 / 3.14159265359) + 24)) * 60)
```

## Model
### Regression (Model B)

```{r}
# Build models
m1 <- lm(hair_time ~ sex + bs(age) + employed, data = ht2)
jtools::summ(m1, digits = 3, confint = T)
```

#### Conversion of Est. to minutes
```{r}
  print(0.564 * 60)
  minutes <- 0.564 * 60  # Convert hours to minutes
  seconds <- (minutes - floor(minutes)) * 60  # Extract remaining seconds
sprintf("%d min %d sec", floor(minutes), round(seconds))

```

### Model assumptions
```{r, fig.width=8, fig.height=8}
suppressMessages(library(performance))
check_model(m1)

```

### Plot (estimated marginal means) 

```{r}
# Compute emmeans for pairwise comparisons
emmeans_results <- emmeans(m1, ~ employed)
pairwise_comparison <- contrast(emmeans_results, method = "pairwise")

# Extract p-value
p_value_work <- summary(pairwise_comparison)$p.value

# Compute t-test for hair_time between employment groups
t_test_result <- t.test(ht2$hair_time ~ ht2$employed, var.equal = TRUE)
p_value_violin <- t_test_result$p.value

# Get effect estimates
eff_empl <- ggeffects::predict_response(m1, margin = "marginalmeans", terms = "employed")

# Create mod_work plot
mod_work <- ggplot() +
  geom_point(data = eff_empl, aes(x = x,  y = predicted, color = x), size = 3) +
  geom_errorbar(data = eff_empl, aes(x = x, ymin = conf.low, ymax = conf.high, color = x),
                linewidth = .4, width = .05) + 
  theme_bw() +
  xlab("Working") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-3.75, -3.5, -3.25, -3, -2.75, -2.5),
                     labels = c("20.25", "20.50", "20.75", "21.00", "21.25", "21.50"),
                     limits = c(-3.8, -2.7)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line =  element_line(linewidth = .4),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  scale_color_manual(values = c("#6D93B0", "#B08A6D"), name = "Working") +
  scale_fill_manual(values = c("#6D93B0", "#B08A6D"), name = "Working") +
  stat_pvalue_manual(
    data = data.frame(x = "Yes", y = 4.5, group1 = "No", group2 = "Yes",
                      label = sprintf("p = %.3f", p_value_work)),
    y.position = -2.75,
    size = 6
  )
mod_work




```

### Sensitivity analysis (circularity of pDLMO): hair time using sine and cosine components 

```{r}
#rm(list = ls(all = TRUE)) 
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")

library(dplyr)
library(ggplot2)
library(splines)
library(ggeffects)
library(ggpubr)
library(emmeans)

# Filter NA and adjust hair_time
ht2 <- ht2 %>% filter(!is.na(employed))
ht2$hair_time <- ifelse(ht2$hair_time > 13, ht2$hair_time -24, ht2$hair_time)

# 1. Prepare Circular Data
ht2$theta <- (ht2$hair_time %% 24) * (2*pi/24)
ht2$sin_t <- sin(ht2$theta)
ht2$cos_t <- cos(ht2$theta)

# 2. Fit the Two Component Models
m_sin <- lm(sin_t ~ bs(age) + sex + employed, data = ht2)
m_cos <- lm(cos_t ~ bs(age) + sex + employed, data = ht2)

# Get predictions for employed
eff_sin_empl <- predict_response(m_sin, terms = "employed", margin = "marginalmeans")
eff_cos_empl <- predict_response(m_cos, terms = "employed", margin = "marginalmeans")

# Helper function to approximate SE from 95% CI
get_se <- function(fit, upper) (upper - fit) / 1.96

# Create simulation dataframe
sim_data_empl <- data.frame(
  employed = eff_sin_empl$x,
  sin_fit = eff_sin_empl$predicted,
  sin_se  = get_se(eff_sin_empl$predicted, eff_sin_empl$conf.high),
  cos_fit = eff_cos_empl$predicted,
  cos_se  = get_se(eff_cos_empl$predicted, eff_cos_empl$conf.high)
)

# Run simulation
n_sims <- 1000
results_empl <- list()

for(i in 1:nrow(sim_data_empl)) {
  s_sim <- rnorm(n_sims, sim_data_empl$sin_fit[i], sim_data_empl$sin_se[i])
  c_sim <- rnorm(n_sims, sim_data_empl$cos_fit[i], sim_data_empl$cos_se[i])
  
  angles <- atan2(s_sim, c_sim)
  times  <- (angles %% (2*pi)) * 24/(2*pi)
  times_shifted <- ifelse(times < 12, times + 24, times)
  
  results_empl[[i]] <- data.frame(
    x = sim_data_empl$employed[i],
    hair_time = mean(times_shifted),
    conf.low  = quantile(times_shifted, 0.025),
    conf.high = quantile(times_shifted, 0.975)
  )
}

eff_empl_final <- do.call(rbind, results_empl)

# Compute p-value for circular model (employment effect)
emmeans_sin <- emmeans(m_sin, ~ employed)
emmeans_cos <- emmeans(m_cos, ~ employed)

pairwise_sin <- contrast(emmeans_sin, method = "pairwise")
pairwise_cos <- contrast(emmeans_cos, method = "pairwise")

p_value_sin <- summary(pairwise_sin)$p.value
p_value_cos <- summary(pairwise_cos)$p.value
p_value_circular_empl <- max(p_value_sin, p_value_cos)

print(paste("Circular model - Sin component p-value:", p_value_sin))
print(paste("Circular model - Cos component p-value:", p_value_cos))
print(paste("Circular model - Combined p-value:", p_value_circular_empl))

# Create plot
mod_work_circular <- ggplot() +
  geom_point(data = eff_empl_final, aes(x = x, y = hair_time, color = x), size = 3) +
  geom_errorbar(data = eff_empl_final, aes(x = x, ymin = conf.low, ymax = conf.high, color = x),
                linewidth = .4, width = .05) + 
  theme_bw() +
  xlab("Working") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(20, 20.25, 20.5, 20.75, 21, 21.25),
                     labels = c("20:00", "20.25", "20.50", "20.75", "21", "21.25")) +
  coord_cartesian(ylim = c(20, 21.25)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line =  element_line(linewidth = .4),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  scale_color_manual(values = c("#6D93B0", "#B08A6D"), name = "Working") +
  scale_fill_manual(values = c("#6D93B0", "#B08A6D"), name = "Working") +
  stat_pvalue_manual(
    data = data.frame(x = "Yes", y = 4.5, group1 = "No", group2 = "Yes",
                      label = sprintf("p = %.3f", p_value_circular_empl)),
    y.position = 21.15,
    size = 6
  )

mod_work_circular
```

### Model B comparison: Spline-based vs Circular pDLMO

```{r}
# ============================================================================
# COMPARING SPLINE-BASED MODEL vs CIRCULAR MODEL - MODEL (3)
# ============================================================================
# For Model (3): hair_time ~ sex + bs(age) + employed
# Focus: Employment effect with age and sex adjustment
# ============================================================================

cat("\n=== MODEL (3) COMPARISON: SPLINE-BASED vs CIRCULAR ===\n\n")

# Spline-based Model (m1)
cat("SPLINE-BASED MODEL (bs(age)):\n")
cat("  Formula: hair_time ~ sex + bs(age) + employed\n")
cat("  AIC:", AIC(m1), "\n")
cat("  BIC:", BIC(m1), "\n")
cat("  Log-Likelihood:", logLik(m1)[1], "\n")
cat("  R-squared:", summary(m1)$r.squared, "\n")
cat("  Adj R-squared:", summary(m1)$adj.r.squared, "\n")
cat("  N parameters:", length(coef(m1)), "\n\n")

# Circular Model (m_sin and m_cos)
cat("CIRCULAR MODEL:\n")
cat("  Formula (Sin): sin_t ~ sex + bs(age) + employed\n")
cat("  Formula (Cos): cos_t ~ sex + bs(age) + employed\n")
cat("  AIC (Sin):", AIC(m_sin), "\n")
cat("  AIC (Cos):", AIC(m_cos), "\n")
cat("  AIC (Combined):", AIC(m_sin) + AIC(m_cos), "\n")
cat("  BIC (Sin):", BIC(m_sin), "\n")
cat("  BIC (Cos):", BIC(m_cos), "\n")
cat("  BIC (Combined):", BIC(m_sin) + BIC(m_cos), "\n")
cat("  Log-Likelihood (Combined):", logLik(m_sin)[1] + logLik(m_cos)[1], "\n")
cat("  R-squared (Sin):", summary(m_sin)$r.squared, "\n")
cat("  R-squared (Cos):", summary(m_cos)$r.squared, "\n")
cat("  N parameters (total):", length(coef(m_sin)) + length(coef(m_cos)), "\n\n")

# Comparison
cat("COMPARISON:\n")
delta_aic <- (AIC(m_sin) + AIC(m_cos)) - AIC(m1)
delta_bic <- (BIC(m_sin) + BIC(m_cos)) - BIC(m1)
cat("  Delta AIC (Circular - Linear):", delta_aic, "\n")
cat("  Delta BIC (Circular - Linear):", delta_bic, "\n\n")

if(delta_aic < -2) {
  cat("  CONCLUSION (AIC): Circular model is BETTER (Delta AIC < -2)\n")
} else if(delta_aic > 2) {
  cat("  CONCLUSION (AIC): Linear model is BETTER (Delta AIC > 2)\n")
} else {
  cat("  CONCLUSION (AIC): Models are equivalent (|Delta AIC| <= 2)\n")
}

if(delta_bic < 0) {
  cat("  CONCLUSION (BIC): Circular model is BETTER (Delta BIC < 0)\n")
} else {
  cat("  CONCLUSION (BIC): Linear model is BETTER (Delta BIC > 0)\n")
}

cat("\n  NOTE: Lower AIC/BIC indicates better model fit.\n")
cat("  The circular model accounts for the periodic nature of time data.\n")
cat("\n=============================================================\n")
```


# Figure 5B

## Distribution

```{r, warning=FALSE}
rm(list = ls(all = TRUE)) 

library(splines)

load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht4.Rda")

ht4$hair_time <- ifelse(ht4$hair_time > 13, ht4$hair_time -24, ht4$hair_time)


# plot ----------------------------------------------------

mean = mean(ht4$hair_time)
median = median(ht4$hair_time)
print(paste0("Mean of hairtime: ", mean(ht4$hair_time)))
print(paste0("Median of hairtime: ", median(ht4$hair_time)))
distr4 <- 
  ggplot(ht4) +
  geom_histogram(data = ht4, aes(x = hair_time),
                 fill = "grey", color = "black", 
                 breaks = seq(-12,12,.5),
                 alpha = .5, linewidth = .1) +
  theme_bw() +
  scale_y_continuous(name = "Distribution (counts)", 
                     breaks = seq(0, 500, 100),
                     limits = c(0, 350)) +
  scale_x_continuous(limits = c(-12,12),
                     breaks = seq(-12,12,2),
                     labels = c("12:00","14:00", "16:00",
                                "18:00","20:00", "22:00",
                                "00:00",
                                "02:00", "04:00", "06:00",
                                "08:00", "10:00", "12:00"),
                     name = "Predicted DLMO (hours)") +
  geom_vline(xintercept = mean, color = "red", linetype = "dashed") +
  geom_vline(xintercept = median, color = "blue", linetype = "dashed") +
  # add annotation for vlines
  annotate("text", x = mean+1, y = 800, label = "Mean", color = "red", size = 3) +
  annotate("text", x = median-1, y = 800, label = "Median", color = "blue", size = 3) +
  theme_bw()+
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        legend.position = "none",
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),  # Add this line
        axis.line =  element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -3),
        axis.title.y = element_text(vjust = 3)) 
distr4
```

## Comparing groups

### Testing

```{r}
suppressMessages(library(car))
leveneTest(hair_time ~ we_wd, data = ht4)
# p > 0.05 → variances are similar → you can use var.equal = TRUE
# p ≤ 0.05 → variances differ → use var.equal = FALSE (Welch)
# Perform t-test
t.test(ht4$hair_time~ht4$we_wd, var.equal = T) 
wilcox.test(ht4$hair_time~ht4$we_wd) # sign
```
### Plot
```{r}
# Create violin plot
# Compute t-test for hair_time between weekday groups
t_test_result <- t.test(ht4$hair_time ~ ht4$we_wd, var.equal = TRUE)
p_value_violin <- t_test_result$p.value

violin <- ggplot(ht4, aes(x = we_wd, y = hair_time)) + 
  geom_violin(aes(color = we_wd, fill = we_wd)) +
  geom_boxplot(color = "black", alpha = .5, width = .5) +
  theme(legend.position = "none") +
  scale_color_manual(values = c("#6D93B0", "#B08A6D")) +
  scale_fill_manual(values = c("#6D93B0", "#B08A6D")) +
  xlab("Sampling weekday") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-10, -7.5, -5, -2.5, 0, 2.5, 5.0),
                     labels = c("14.00", "16.50", "19.00", "21.50", "00.00", "02.50", "05.00"),
                     limits = c(-10, 5.1))  +
  theme_bw() +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line =  element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  stat_pvalue_manual(
    data = data.frame(group1 = "Wed-Fri", group2 = "Sun",
                      label = sprintf("p = %.3f", p_value_violin)),
    y.position = 5,
    size = 4
  )
violin
```


### Sensitivity analysis (circularity of pDLMO): Watson-Williams test (circular) 

```{r}
suppressMessages(library(circular))
ht4$we_wd <- factor(ht4$we_wd, levels = c("wd", "we"), labels = c("Wed-Fri", "Sun"))

# Extract numeric vectors for M and F
wd <- ht4 %>% filter(we_wd == "Wed-Fri") %>% pull(hair_time)
we <- ht4 %>% filter(we_wd == "Sun") %>% pull(hair_time)

# Perform t-test (check if ok)
print(t.test(wd, we))

# Convert hours to radians (since 24 hours = 2π radians)
wd_rad <- wd * (2 * pi / 24)
we_rad <- we * (2 * pi / 24)
 
# Ensure all values are within [0, 2π]
wd_rad <- (wd_rad + 2 * pi) %% (2 * pi)
we_rad <- (we_rad + 2 * pi) %% (2 * pi)
 
# Convert to circular data
wd_circ <- circular(wd_rad, units = "radians")
we_circ <- circular(we_rad, units = "radians")
 
# Perform Watson-Williams test
watson.williams.test(list(wd_circ, we_circ))
 
# print((-0.8551704 * 12 / 3.14159265359) + 24)
# print((-0.9232520 * 12 / 3.14159265359) + 24)
#  
# print((((-0.8551704 * 12 / 3.14159265359) + 24)-((-0.9232520 * 12 / 3.14159265359) + 24)) * 60)

```

## Model

### Regression (Model C)

```{r}
m1 <- lm(hair_time ~  sex + bs(age) + we_wd, data = ht4) 
jtools::summ(m1, digits = 3, confint = T)


# get effects ----------------------------------------------------

eff_wdwe <- ggeffects::predict_response(m1, margin = "marginalmeans", terms = "we_wd")
# -------------------------------------------------------------- # 

```

#### Conversion of Est. to minutes

```{r}
  0.208 * 60 
  minutes <- 0.208 * 60  # Convert hours to minutes
  seconds <- (minutes - floor(minutes)) * 60  # Extract remaining seconds
  sprintf("%d min %d sec", floor(minutes), round(seconds))
```

### Model assumptions
```{r, fig.width=8, fig.height=8}
suppressMessages(library(performance))
check_model(m1)

```

### Plot (estimated marginal means) 

```{r}
library(emmeans)
# Compute emmeans for pairwise comparison
emmeans_results <- emmeans(m1, ~ we_wd)
pairwise_comparison <- contrast(emmeans_results, method = "pairwise")

# Extract p-value
p_value_weekday <- summary(pairwise_comparison)$p.value

# Compute t-test for hair_time between weekday groups
t_test_result <- t.test(ht4$hair_time ~ ht4$we_wd, var.equal = TRUE)
p_value_violin <- t_test_result$p.value

# Get effect estimates
eff_wdwe <- ggeffects::predict_response(m1, margin = "marginalmeans", terms = "we_wd")

# Create mod_sampling_date plot
mod_sampling_date <- ggplot() +
  geom_point(data = eff_wdwe, aes(x = x, y = predicted, color = x), size = 3) +
  geom_errorbar(data = eff_wdwe, aes(x = x, ymin = conf.low, ymax = conf.high, color = x),
                linewidth = .4, width = .05) + 
  theme_bw() +
  xlab("Sampling date") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-4.0, -3.75, -3.5, -3.25, -3.0),
                     labels = c("20.00", "20.25", "20.50", "20.75", "21.00"),
                     limits = c(-4.0, -2.95)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line =  element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  scale_color_manual(values = c("#6D93B0", "#B08A6D"), name = "Sampling date") +
  scale_fill_manual(values = c("#6D93B0", "#B08A6D"), name = "") +
  stat_pvalue_manual(
    data = data.frame(x = "Wed-Fri", y = 4, group1 = "Sun", group2 = "Wed-Fri",
                      label = sprintf("p = %.3f", p_value_weekday)),
    y.position = -3.25,
    size = 6
  )


mod_sampling_date


```

### Sensitivity analysis (circularity of pDLMO): hair time using sine and cosine components 

```{r}
#rm(list = ls(all = TRUE))
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht4.Rda")

library(dplyr)
library(ggplot2)
library(splines)
library(ggeffects)
library(ggpubr)
library(emmeans)

# Adjust hair_time and factor
ht4$hair_time <- ifelse(ht4$hair_time > 13, ht4$hair_time -24, ht4$hair_time)
ht4$we_wd <- factor(ht4$we_wd, levels = c("wd", "we"), labels = c("Wed-Fri", "Sun"))

# 1. Prepare Circular Data
ht4$theta <- (ht4$hair_time %% 24) * (2*pi/24)
ht4$sin_t <- sin(ht4$theta)
ht4$cos_t <- cos(ht4$theta)

# 2. Fit the Two Component Models
m_sin <- lm(sin_t ~ bs(age) + sex + we_wd, data = ht4)
m_cos <- lm(cos_t ~ bs(age) + sex + we_wd, data = ht4)

# Get predictions for we_wd
eff_sin_wd <- predict_response(m_sin, terms = "we_wd", margin = "marginalmeans")
eff_cos_wd <- predict_response(m_cos, terms = "we_wd", margin = "marginalmeans")

# Helper function to approximate SE from 95% CI
get_se <- function(fit, upper) (upper - fit) / 1.96

# Create simulation dataframe
sim_data_wd <- data.frame(
  we_wd = eff_sin_wd$x,
  sin_fit = eff_sin_wd$predicted,
  sin_se  = get_se(eff_sin_wd$predicted, eff_sin_wd$conf.high),
  cos_fit = eff_cos_wd$predicted,
  cos_se  = get_se(eff_cos_wd$predicted, eff_cos_wd$conf.high)
)

# Run simulation
n_sims <- 1000
results_wd <- list()

for(i in 1:nrow(sim_data_wd)) {
  s_sim <- rnorm(n_sims, sim_data_wd$sin_fit[i], sim_data_wd$sin_se[i])
  c_sim <- rnorm(n_sims, sim_data_wd$cos_fit[i], sim_data_wd$cos_se[i])
  
  angles <- atan2(s_sim, c_sim)
  times  <- (angles %% (2*pi)) * 24/(2*pi)
  times_shifted <- ifelse(times < 12, times + 24, times)
  
  results_wd[[i]] <- data.frame(
    x = sim_data_wd$we_wd[i],
    hair_time = mean(times_shifted),
    conf.low  = quantile(times_shifted, 0.025),
    conf.high = quantile(times_shifted, 0.975)
  )
}

eff_wdwe_final <- do.call(rbind, results_wd)

# Compute p-value for circular model (weekday effect)
emmeans_sin <- emmeans(m_sin, ~ we_wd)
emmeans_cos <- emmeans(m_cos, ~ we_wd)

pairwise_sin <- contrast(emmeans_sin, method = "pairwise")
pairwise_cos <- contrast(emmeans_cos, method = "pairwise")

p_value_sin <- summary(pairwise_sin)$p.value
p_value_cos <- summary(pairwise_cos)$p.value
p_value_circular_wd <- max(p_value_sin, p_value_cos)

print(paste("Circular model - Sin component p-value:", p_value_sin))
print(paste("Circular model - Cos component p-value:", p_value_cos))
print(paste("Circular model - Combined p-value:", p_value_circular_wd))

# Create plot
mod_sampling_date_circular <- ggplot() +
  geom_point(data = eff_wdwe_final, aes(x = x, y = hair_time, color = x), size = 3) +
  geom_errorbar(data = eff_wdwe_final, aes(x = x, ymin = conf.low, ymax = conf.high, color = x),
                linewidth = .4, width = .05) + 
  theme_bw() +
  xlab("Sampling date") +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(20, 20.25, 20.5, 20.75, 21),
                     labels = c("20.00", "20.25", "20.50", "20.75", "21.00")) +
  coord_cartesian(ylim = c(20, 21)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(.8, .8, .8, .6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "none",
        axis.line =  element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = .1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3)) +
  scale_color_manual(values = c("#6D93B0", "#B08A6D"), name = "Sampling date") +
  scale_fill_manual(values = c("#6D93B0", "#B08A6D"), name = "Sampling date") +
  stat_pvalue_manual(
    data = data.frame(x = "Wed-Fri", y = 4, group1 = "Sun", group2 = "Wed-Fri",
                      label = sprintf("p = %.3f", p_value_circular_wd)),
    y.position = 20.75,
    size = 6
  )

mod_sampling_date_circular
```

### Model C comparison: Spline-based vs Circular pDLMO

```{r}
# ============================================================================
# COMPARING SPLINE-BASED MODEL vs CIRCULAR MODEL - MODEL (4)
# ============================================================================
# For Model (4): hair_time ~ sex + bs(age) + we_wd
# Focus: Weekday/Weekend effect with age and sex adjustment
# ============================================================================

cat("\n=== MODEL (4) COMPARISON: SPLINE-BASED vs CIRCULAR ===\n\n")

# Spline-based Model (m1)
cat("SPLINE-BASED MODEL (bs(age)):\n")
cat("  Formula: hair_time ~ sex + bs(age) + we_wd\n")
cat("  AIC:", AIC(m1), "\n")
cat("  BIC:", BIC(m1), "\n")
cat("  Log-Likelihood:", logLik(m1)[1], "\n")
cat("  R-squared:", summary(m1)$r.squared, "\n")
cat("  Adj R-squared:", summary(m1)$adj.r.squared, "\n")
cat("  N parameters:", length(coef(m1)), "\n\n")

# Circular Model (m_sin and m_cos)
cat("CIRCULAR MODEL:\n")
cat("  Formula (Sin): sin_t ~ sex + bs(age) + we_wd\n")
cat("  Formula (Cos): cos_t ~ sex + bs(age) + we_wd\n")
cat("  AIC (Sin):", AIC(m_sin), "\n")
cat("  AIC (Cos):", AIC(m_cos), "\n")
cat("  AIC (Combined):", AIC(m_sin) + AIC(m_cos), "\n")
cat("  BIC (Sin):", BIC(m_sin), "\n")
cat("  BIC (Cos):", BIC(m_cos), "\n")
cat("  BIC (Combined):", BIC(m_sin) + BIC(m_cos), "\n")
cat("  Log-Likelihood (Combined):", logLik(m_sin)[1] + logLik(m_cos)[1], "\n")
cat("  R-squared (Sin):", summary(m_sin)$r.squared, "\n")
cat("  R-squared (Cos):", summary(m_cos)$r.squared, "\n")
cat("  N parameters (total):", length(coef(m_sin)) + length(coef(m_cos)), "\n\n")

# Comparison
cat("COMPARISON:\n")
delta_aic <- (AIC(m_sin) + AIC(m_cos)) - AIC(m1)
delta_bic <- (BIC(m_sin) + BIC(m_cos)) - BIC(m1)
cat("  Delta AIC (Circular - Linear):", delta_aic, "\n")
cat("  Delta BIC (Circular - Linear):", delta_bic, "\n\n")

if(delta_aic < -2) {
  cat("  CONCLUSION (AIC): Circular model is BETTER (Delta AIC < -2)\n")
} else if(delta_aic > 2) {
  cat("  CONCLUSION (AIC): Linear model is BETTER (Delta AIC > 2)\n")
} else {
  cat("  CONCLUSION (AIC): Models are equivalent (|Delta AIC| <= 2)\n")
}

if(delta_bic < 0) {
  cat("  CONCLUSION (BIC): Circular model is BETTER (Delta BIC < 0)\n")
} else {
  cat("  CONCLUSION (BIC): Linear model is BETTER (Delta BIC > 0)\n")
}

cat("\n  NOTE: Lower AIC/BIC indicates better model fit.\n")
cat("  The circular model accounts for the periodic nature of time data.\n")
cat("\n=============================================================\n")
```


# Figure 5C

## Distribution

```{r}
# Load necessary libraries
library(ggplot2)
library(tidyverse)
library(dplyr)
library(splines)
library(ggpubr)
library(ggeffects)
library(plotrix)
# Clear environment
rm(list = ls(all = TRUE))

# Load dataset and filter
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")
# replace person with ht2$age = 51.5 to 52
ht2$age[ht2$age == 51.5] <- 52

# Adjust hair_time (for values > 13)
ht2$hair_time <- ifelse(ht2$hair_time > 13, ht2$hair_time - 24, ht2$hair_time)

#---------------------- Distribution Plot ----------------------#
mean_val <- mean(ht2$hair_time)
median_val <- median(ht2$hair_time)
print(paste0("Mean of hairtime: ", mean(ht2$hair_time)))
print(paste0("Median of hairtime: ", median(ht2$hair_time)))

distr2 <- ggplot(ht2) +
  geom_histogram(aes(x = hair_time),
                 fill = "grey", color = "black", 
                 breaks = seq(-12, 12, 0.5),
                 alpha = 0.5, linewidth = 0.1) +
  theme_bw() +
  scale_y_continuous(name = "Distribution (counts)", breaks = seq(0, 500, 100)) +
  scale_x_continuous(limits = c(-12, 12),
                     breaks = seq(-12, 12, 2),
                     labels = c("12:00","14:00", "16:00",
                                "18:00","20:00", "22:00",
                                "00:00","02:00", "04:00", "06:00",
                                "08:00", "10:00", "12:00"),
                     name = "Predicted DLMO (hours)") +
  geom_vline(xintercept = mean_val, color = "red", linetype = "dashed") +
  geom_vline(xintercept = median_val, color = "blue", linetype = "dashed") +
  annotate("text", x = mean_val + 1, y = 800, label = "Mean", color = "red", size = 3) +
  annotate("text", x = median_val - 1, y = 800, label = "Median", color = "blue", size = 3) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        legend.position = "none",
        plot.margin = margin(0.8, 0.8, 0.8, 0.6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.line = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = 0.1),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -3),
        axis.title.y = element_text(vjust = 3))
distr2
```
## Association between variables 

### Tests

```{r}
suppressMessages(library(circular))
suppressMessages(library(CircStats))
suppressMessages(library(dplyr))
#suppressMessages(library(Directional))


#plot(ht2$workdays, ht2$hair_time)
#abline(lm(hair_time ~ workdays, data = ht2), col = "red")

shapiro.test(ht2$workdays)
shapiro.test(ht2$hair_time)

#boxplot(ht2$workdays)
#boxplot(ht2$hair_time)

cor.test(ht2$workdays, ht2$hair_time, method="pearson")
cor.test(ht2$workdays, ht2$hair_time, method="spearman")

normal1 <- shapiro.test(ht2$workdays)$p.value > 0.05
normal2 <- shapiro.test(ht2$hair_time)$p.value > 0.05

if(normal1 & normal2){
  cor.test(ht2$workdays, ht2$hair_time, method="pearson")
} else {
  cor.test(ht2$workdays, ht2$hair_time, method="spearman")
}


# Spearman correlation test
cor.test(ht2$workdays, ht2$hair_time, method = "spearman")
# Linear regression
model <- lm(hair_time ~ workdays, data = ht2)
summary(model)
```

### Sensitivity analysis (circularity of pDLMO): Watson-Williams test (circular)
```{r}
all <- ht2  %>% pull(hair_time)

# Convert hours to radians (since 24 hours = 2π radians)
all_rad <- all * (2 * pi / 24)

# Ensure all values are within [0, 2π]
all_rad <- (all_rad + 2 * pi) %% (2 * pi)

# Convert to circular data
all_circ <- circular(all_rad, units = "radians")

# 1. Linear circular correlation
circlin.cor(theta = all_circ, x = ht2$workdays, rads = TRUE)

```

## Models

 ### Regression (model D)

```{r}
# Fit model (using workdays as predictor)
m1 <- lm(hair_time ~ sex + bs(age) + bs(workdays), data = ht2)
m2 <- lm(hair_time ~ sex + bs(age) + workdays, data = ht2)
jtools::summ(m2, digits = 3, confint = T)
jtools::summ(m1, digits = 3, confint = T)
```

#### BIC: 

```{r}
print(paste0("BIC for model 1 (workdays splines): ", BIC(m1)))
print(paste0("BIC for model 2 (workdays linear): ", BIC(m2)))
print("Selected: model 2")
```

#### Conversion of Est. to minutes

```{r}
0.121 * 60
  minutes <- 0.121 * 60  # Convert hours to minutes
  seconds <- (minutes - floor(minutes)) * 60  # Extract remaining seconds
sprintf("%d min %d sec", floor(minutes), round(seconds))

```

### Model assumptions

```{r, fig.width=8, fig.height=8}
suppressMessages(library(performance))
check_model(m2)
```

### Plot
 
```{r, warning=FALSE}
# Predicted effects for workdays
eff_wd <- ggeffects::predict_response(m2, margin = "marginalmeans", terms = "workdays")
eff_wd <- eff_wd %>% rename(hair_time = predicted)

# Compute means by workdays (for raw hair_time)
means <- ht2 %>% group_by(workdays) %>% 
  summarise(mean_age = mean(age),
            se_hair_time = std.error(hair_time),
            mean_hair_time = mean(hair_time),
            n = n())
means$x <- as.numeric(as.character(means$workdays))

# For adjusted dots: correct hair_time using model predictions
wd_age <- ggeffects::predict_response(m2, margin = "marginalmeans", terms = "age[18:70]")
wd_all <- ggeffects::predict_response(m2, margin = "marginalmeans", terms = c("age[18:70]", "sex"))
wd_all <- as.data.frame(wd_all)
corr_wd <- wd_all %>% dplyr::select(x, group, predicted)
colnames(corr_wd) <- c("age", "sex", "predicted_hair_time")

db <- merge(ht2, corr_wd, by = c("sex", "age"), all.x = TRUE)
db$hair_time_corrected <- db$hair_time - db$predicted_hair_time +
  wd_age$predicted[wd_age$x == round(mean(db$age))]

means2 <- db %>% group_by(workdays) %>% 
  summarise(mean_age = mean(age),
            se_hair_time = std.error(hair_time_corrected),
            mean_hair_time = mean(hair_time_corrected),
            n = n())
means2$x <- as.numeric(as.character(means2$workdays))

# Build workdays figure
workdays_fig <- ggplot() +
  geom_point(data = means, aes(x = x, y = mean_hair_time),
             color = "#a3bbbf", alpha = 0.5) +
  geom_errorbar(data = means, aes(x = x, ymin = mean_hair_time - se_hair_time, ymax = mean_hair_time + se_hair_time),
                linewidth = 0.3, width = 0.1, color = "#a3bbbf", alpha = 0.5) +
  geom_point(data = means2, aes(x = x, y = mean_hair_time)) +
  geom_errorbar(data = means2, aes(x = x, ymin = mean_hair_time - se_hair_time, ymax = mean_hair_time + se_hair_time),
                linewidth = 0.3, width = 0.1) +
  geom_text(data = means2, aes(label = n, x = workdays + 0.25, y = mean_hair_time + 0.35),
            color = "#ff4d00") +
  geom_line(data = eff_wd, aes(x = x, y = hair_time)) +
  geom_ribbon(data = eff_wd, aes(x = x, ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  theme_bw() +
  scale_x_continuous("Workdays", breaks = seq(0, 7, 1)) +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(-4, -3.5, -3, -2.5, -2, -1.5, -1, -0.5, 0),
                     labels = c("20.00", "20.50", "21.00",
                                "21.50", "22.00", "22.50",
                                "23.00", "23.50", "00.00"),
                     limits = c(-4, -0.6)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(0.8, 0.8, 0.8, 0.6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.line = element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = 0.1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3),
        legend.position = "bottom")
workdays_fig

```

### Model D comparison: Linear vs Circular pDLMO

```{r, warning=FALSE}
#rm(list = ls(all = TRUE))
load("/Volumes/ALI_1TB/Projects/BodyTime/BodyTime_Final_Version/06_FinalDataset_2024/ht2.Rda")

library(dplyr)
library(ggplot2)
library(splines)
library(ggeffects)
library(plotrix)

# Replace person with age = 51.5 to 52
ht2$age[ht2$age == 51.5] <- 52
# Adjust hair_time
ht2$hair_time <- ifelse(ht2$hair_time > 13, ht2$hair_time - 24, ht2$hair_time)

# 1. Prepare Circular Data
ht2$theta <- (ht2$hair_time %% 24) * (2*pi/24)
ht2$sin_t <- sin(ht2$theta)
ht2$cos_t <- cos(ht2$theta)

# 2. Fit the Two Component Models
m_sin <- lm(sin_t ~ bs(age) + sex + workdays, data = ht2)
m_cos <- lm(cos_t ~ bs(age) + sex + workdays, data = ht2)

# Get predictions for workdays
eff_sin_wd <- predict_response(m_sin, terms = "workdays", margin = "marginalmeans")
eff_cos_wd <- predict_response(m_cos, terms = "workdays", margin = "marginalmeans")

# Helper function to approximate SE from 95% CI
get_se <- function(fit, upper) (upper - fit) / 1.96

# Create simulation dataframe for each workday value
sim_data_wd <- data.frame(
  workdays = eff_sin_wd$x,
  sin_fit = eff_sin_wd$predicted,
  sin_se  = get_se(eff_sin_wd$predicted, eff_sin_wd$conf.high),
  cos_fit = eff_cos_wd$predicted,
  cos_se  = get_se(eff_cos_wd$predicted, eff_cos_wd$conf.high)
)

# Run simulation
n_sims <- 1000
results_wd <- list()

for(i in 1:nrow(sim_data_wd)) {
  s_sim <- rnorm(n_sims, sim_data_wd$sin_fit[i], sim_data_wd$sin_se[i])
  c_sim <- rnorm(n_sims, sim_data_wd$cos_fit[i], sim_data_wd$cos_se[i])
  
  angles <- atan2(s_sim, c_sim)
  times  <- (angles %% (2*pi)) * 24/(2*pi)
  times_shifted <- ifelse(times < 12, times + 24, times)
  
  results_wd[[i]] <- data.frame(
    x = sim_data_wd$workdays[i],
    hair_time = mean(times_shifted),
    conf.low  = quantile(times_shifted, 0.025),
    conf.high = quantile(times_shifted, 0.975)
  )
}

eff_wd_final <- do.call(rbind, results_wd)

# Prepare data for raw means (age-corrected)
# Get age and sex effects for correction
wd_age <- predict_response(m_sin, terms = "age[18:70]", margin = "marginalmeans")
wd_all_sin <- predict_response(m_sin, terms = c("age[18:70]", "sex"), margin = "marginalmeans")
wd_all_cos <- predict_response(m_cos, terms = c("age[18:70]", "sex"), margin = "marginalmeans")

# For simplicity, calculate raw means per workday with shifted times
ht2$hair_time_shifted <- ifelse(ht2$hair_time < 12, ht2$hair_time + 24, ht2$hair_time)

means <- ht2 %>% group_by(workdays) %>% 
  summarise(mean_age = mean(age),
            se_hair_time = std.error(hair_time_shifted),
            mean_hair_time = mean(hair_time_shifted),
            n = n())
means$x <- as.numeric(as.character(means$workdays))

# Build workdays circular figure
workdays_fig_circular <- ggplot() +
  geom_point(data = means, aes(x = x, y = mean_hair_time),
             color = "#a3bbbf", alpha = 0.5) +
  geom_errorbar(data = means, aes(x = x, ymin = mean_hair_time - se_hair_time, ymax = mean_hair_time + se_hair_time),
                linewidth = 0.3, width = 0.1, color = "#a3bbbf", alpha = 0.5) +
  geom_text(data = means, aes(label = n, x = workdays + 0.25, y = mean_hair_time + 0.35),
            color = "#ff4d00") +
  geom_line(data = eff_wd_final, aes(x = x, y = hair_time), color = "firebrick", linewidth = 1) +
  geom_ribbon(data = eff_wd_final, aes(x = x, ymin = conf.low, ymax = conf.high), 
              fill = "firebrick", alpha = 0.2) +
  theme_bw() +
  scale_x_continuous("Workdays", breaks = seq(0, 7, 1)) +
  scale_y_continuous("Predicted DLMO (hours)",
                     breaks = c(20, 20.5, 21, 21.5, 22, 22.5, 23, 23.5, 24),
                     labels = c("20.00", "20.50", "21.00", "21.50", "22.00", 
                                "22.50", "23.00", "23.50", "00.00")) +
  coord_cartesian(ylim = c(19.9, 23.5)) +
  theme(plot.background = element_rect(fill = "transparent", colour = "transparent"),
        plot.margin = margin(0.8, 0.8, 0.8, 0.6, "cm"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.line = element_line(),
        panel.grid.minor.y = element_blank(),
        panel.grid = element_line(linewidth = 0.1),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.x = element_text(vjust = -2),
        axis.title.y = element_text(vjust = 3),
        legend.position = "bottom")
workdays_fig_circular

```

### Model D comparison: Spline-based vs Circular

```{r}
# ============================================================================
# COMPARING SPLINE-BASED MODEL vs CIRCULAR MODEL - MODEL (5)
# ============================================================================
# For Model (5): hair_time ~ sex + bs(age) + workdays
# Focus: Workdays (continuous) effect with age and sex adjustment
# ============================================================================

cat("\n=== MODEL (5) COMPARISON: SPLINE-BASED vs CIRCULAR ===\n\n")

# Spline-based Model (m2 is the selected spline-based model)
cat("SPLINE-BASED MODEL (bs(age)):\n")
cat("  Formula: hair_time ~ sex + bs(age) + workdays\n")
cat("  AIC:", AIC(m2), "\n")
cat("  BIC:", BIC(m2), "\n")
cat("  Log-Likelihood:", logLik(m2)[1], "\n")
cat("  R-squared:", summary(m2)$r.squared, "\n")
cat("  Adj R-squared:", summary(m2)$adj.r.squared, "\n")
cat("  N parameters:", length(coef(m2)), "\n\n")

# Circular Model (m_sin and m_cos)
cat("CIRCULAR MODEL:\n")
cat("  Formula (Sin): sin_t ~ sex + bs(age) + workdays\n")
cat("  Formula (Cos): cos_t ~ sex + bs(age) + workdays\n")
cat("  AIC (Sin):", AIC(m_sin), "\n")
cat("  AIC (Cos):", AIC(m_cos), "\n")
cat("  AIC (Combined):", AIC(m_sin) + AIC(m_cos), "\n")
cat("  BIC (Sin):", BIC(m_sin), "\n")
cat("  BIC (Cos):", BIC(m_cos), "\n")
cat("  BIC (Combined):", BIC(m_sin) + BIC(m_cos), "\n")
cat("  Log-Likelihood (Combined):", logLik(m_sin)[1] + logLik(m_cos)[1], "\n")
cat("  R-squared (Sin):", summary(m_sin)$r.squared, "\n")
cat("  R-squared (Cos):", summary(m_cos)$r.squared, "\n")
cat("  N parameters (total):", length(coef(m_sin)) + length(coef(m_cos)), "\n\n")

# Comparison
cat("COMPARISON:\n")
delta_aic <- (AIC(m_sin) + AIC(m_cos)) - AIC(m2)
delta_bic <- (BIC(m_sin) + BIC(m_cos)) - BIC(m2)
cat("  Delta AIC (Circular - Linear):", delta_aic, "\n")
cat("  Delta BIC (Circular - Linear):", delta_bic, "\n\n")

if(delta_aic < -2) {
  cat("  CONCLUSION (AIC): Circular model is BETTER (Delta AIC < -2)\n")
} else if(delta_aic > 2) {
  cat("  CONCLUSION (AIC): Linear model is BETTER (Delta AIC > 2)\n")
} else {
  cat("  CONCLUSION (AIC): Models are equivalent (|Delta AIC| <= 2)\n")
}

if(delta_bic < 0) {
  cat("  CONCLUSION (BIC): Circular model is BETTER (Delta BIC < 0)\n")
} else {
  cat("  CONCLUSION (BIC): Linear model is BETTER (Delta BIC > 0)\n")
}

cat("\n  NOTE: Lower AIC/BIC indicates better model fit.\n")
cat("  The circular model accounts for the periodic nature of time data.\n")
cat("  Workdays as a continuous predictor may show different patterns in circular space.\n")
cat("\n=============================================================\n")
```